### 2023-10-10 appendices/paper_brief.tex
Updates to discussion

\section{Introduction}
Representation learning has shaped modern computer vision \parencite{simonyan_very_2014} and natural language processing \parencite{devlin_bert_2018}, and more recently speech processing has been subject to the same development \parencite{baevski_wav2vec_2020}. Representation learning has been defined as \emph{``learning representations of the data that make it easier to extract useful information when building classifiers or other predictors"} \parencite{bengio_representation_2013}.  
Unsupervised representation learning is concerned with learning useful representations without the use of human annotations. Usually, a model is first pre-trained on a task where plenty of data is available. The model is then fine-tuned, or used to extract input representations for a smaller model, targeting a task with limited training data. In computer vision, both supervised \parencite{simonyan_very_2014, szegedy_going_2015, he_deep_2015} and unsupervised \parencite{pathak_context_2016, doersch_unsupervised_2015} representation learning have gained attention with supervised representation learning driven by the availability of large annotated datasets \parencite{deng_imagenet_2009}. For text and speech, pre"=training is usually unsupervised as labeled data is difficult to obtain. Although work on text has paved the way, and the two fields share many characteristics, learning representations from speech is a problem faced with a unique set of challenges.

@@

\begin{figure}
    \begin{flushleft}
        \textsc{\small \hspace{0.16\textwidth} self"=supervised \hspace{0.22\textwidth} latent variable}
    \end{flushleft}
    \centering
    \includegraphics[width=0.90\textwidth]{paper_brief/SSL_LVM.pdf}
    \caption[ schematic overview of the self"=supervised and probabilistic latent variable models covered in this survey.]{A schematic overview of the two groups of models covered in this survey. \textit{Left}: A model trained with self"=supervised learning. We take these models to consist of two functions $f(\cdot)$ and $g(\cdot)$ (\cref{ssec:notation}). After pre"=training, $f(\cdot)$ is fine-tuned or used for extracting features $\mathbf{c}$. $g(\cdot)$ is an auxiliary function used to accommodate the self"=supervised pre"=training task. \textit{Right}: A probabilistic latent variable model. In contrast to the self"=supervised model, the functions $f(\cdot)$ and $g(\cdot)$ learn the parameters of distributions $q$ and $p$. The latent variable $\mathbf{z}$ is commonly used for representation learning.}
    \label{fig:ssl_lvm}
\end{figure}

@@

When discussing self"=supervised models, we use $\mathbf{c}_{1:T}$ to denote a contextualized representation. For stochastic latent variable models, we use $\mathbf{z}_{1:T}$ as is customary to the field. 
While some models are frozen and produce representations used as input for downstream tasks (\textbf{\textsc{frz}}, \cref{tab:model-taxonomy}), others are designed to be fine-tuned (\textbf{\textsc{ftn}}, \cref{tab:model-taxonomy}). In either case, we use $f(\cdot)$ to denote the model that is used for the downstream task. We use $g(\cdot)$ to denote any auxiliary model components (e.g., for a reconstruction task we might have $g: \mathbf{c}_t \mapsto \hat{\mathbf{x}}_t$). When a model can be naturally subdivided into multiple components, we simply use $f_*(\cdot)$ where $*$ may be any convenient descriptor. Finally, we often use a subscript when defining a loss, $\mathcal{L}_i$, to imply that the total loss is computed as a sum over $i$.

@@

Combining masking with a contrastive loss, wav2vec 2.0 was the first work to show that a competitive speech recognition model can be learned by fine-tuning a pre-trained model with as little as 10 minutes of labeled data. For this model
%
\begin{align}
    \mathbf{v}_{t} &= f_v(\mathbf{x}_{t-r:t+r}) \\
    \mathbf{c}_{1:T} &= f_c(\mathbf{v}_{1:T} \circ \mathbf{m}_{1:T}) \\
    \mathbf{q}_t &= g_q(\mathbf{v}_t) \enspace . \label{eq_brief: w2v2 qtz}
\end{align}

@@

\begin{sidewaystable*}
    \caption[Classification of selected self"=supervised and probabilistic latent variable models.]{ Selected models classified according to the binary attributes identified throughout the text. The models are sorted according to first publication date on arXiv which might differ from the citation year. \textbf{\textsc{msk}}: masking, \textbf{\textsc{prd}}: prediction, \textbf{\textsc{con}}: contrastive, \textbf{\textsc{rec}}: reconstruction, \textbf{\textsc{qtz}}: quantization, \textbf{\textsc{gen}}: generative, \textbf{\textsc{frz}}: frozen, \textbf{\textsc{ftn}}: fine-tuned, \textbf{\textsc{loc}}: local, \textbf{\textsc{glo}}: global.}
    \label{tab:model-taxonomy}
    \begin{center}
        \renewcommand{\arraystretch}{1.1}
        \resizebox{1\textwidth}{!}{%
        \begin{tabular}{ l l | c c c c c c | c c c | c c } 
            \toprule

            \multicolumn{2}{c}{} & 
            \multicolumn{6}{c}{\footnotesize\textsc{Model and task design}} & 
            \multicolumn{3}{c}{\footnotesize\textsc{Resolution}} &
            \multicolumn{2}{c}{\footnotesize\makebox[0pt][c]{\textsc{Usage}}} \\
            \textbf{\textsc{model}} &
            \textbf{\textsc{pub. date}} &
            \textbf{\textsc{msk}} & 
            \textbf{\textsc{prd}} & 
            \textbf{\textsc{con}} &  
            \textbf{\textsc{rec}} &  
            \textbf{\textsc{qtz}} & 
            \textbf{\textsc{gen}} & 
            \textbf{\textsc{loc}} & 
            \textbf{\textsc{glb}} & 
            \textbf{\textsc{var}} & 
            \textbf{\textsc{frz}} & 
            \textbf{\textsc{ftn}} \\

@@

            \textbf{APC} \footnotesize{\parencite{chung_unsupervised_2019}}          & 2019 Oct. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 5/4
            \textbf{wav2vec} \footnotesize{\parencite{schneider_wav2vec_2019}}       & 2019 Apr. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/4
            \textbf{Mockingjay}  \footnotesize{\parencite{liu_mockingjay_2020}}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark \\ % 25/10-19
            \textbf{wav2vec 2.0} \footnotesize{\parencite{baevski_wav2vec_2020}}     & 2020 Jun. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 20/6
            \textbf{NPC} \footnotesize{\parencite{liu_nonautoregressive_2020}}       & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 1/11
            \textbf{DeCoAR 2.0} \footnotesize{\parencite{ling_decoar_2020}}          & 2020 Dec. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/12
            \textbf{SCPC} \footnotesize{\parencite{bhati_segmental_2021}}            & 2021 Jun. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark \\ % 3/6
            \textbf{HuBERT} \footnotesize{\parencite{hsu_hubert_2021}}               & 2021 Jun. & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 14/6
            \midrule
            \multicolumn{12}{c}{\textsc{Probabilistic latent variable models}} \\
            \midrule
            %                                                           MSK      PRD      CON      REC      QTZ      GEN      LOC      GLO     VAR      FRZ      FTN
            \textbf{VRNN} \footnotesize{\parencite{chung_recurrent_2015}}          & 2015 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
            \textbf{SRNN} \footnotesize{\parencite{fraccaro_sequential_2016}}      & 2016 May  & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
            \textbf{HMM-VAE} \footnotesize{\parencite{ebbers_hidden_2017}}         & 2017 Mar. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
            \textbf{ConvVAE} \footnotesize{\parencite{hsu_learning_2017}}          & 2017 Apr. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
            \textbf{FHVAE} \footnotesize{\parencite{hsu_unsupervised_2017}}        & 2017 Sep. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \cmark & \xmark \\
            \textbf{VQ-VAE} \footnotesize{\parencite{oord_neural_2018}}            & 2017 Nov. & \xmark & \xmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
            \textbf{BHMM-VAE} \footnotesize{\parencite{glarner_full_2018}}         & 2018 Mar. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
            \textbf{STCN} \footnotesize{\parencite{aksan_stcn_2019}}               & 2019 Feb. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
            \textbf{FDMM} \footnotesize{\parencite{khurana_factorial_2019}}        & 2019 Oct. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \cmark & \xmark \\
            \textbf{ConvDMM} \footnotesize{\parencite{khurana_convolutional_2020}} & 2020 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
            \bottomrule
        \end{tabular}
    \end{center}
\end{sidewaystable*}

@@

\textit{Gumbel-softmax approach:} 
Say we want to quantize a vector $\mathbf{v}$ such that it takes one of $K$ possible values. We first map $\mathbf{v}$ to $\mathbf{l} \in \mathbb{R}^K$ and then map $\mathbf{l}$ to a probability vector $\mathbf{p} \in \mathbb{R}^K$ via the Gumbel softmax given by
\begin{align}
    p_i &= \frac{\exp(l_i + n_i) / \tau}{\sum_k^K \exp(l_k + n_k) / \tau} 
\end{align}
for $i=1,\dots,K$. Here $\tau$ is a temperature parameter and $\mathbf{n}\in\mathbb R^K$ is a random vector with $n_i = -\log(-\log(u_i))$ for $u_i \sim \mathcal{U}(0, 1)$.
As  $\tau \rightarrow 0$, $\mathbf{p}$ approaches a one-hot vector. The Gumbel noise $\mathbf{n}$ is a practical way to sample from the untempered categorical distribution (i.e., $\tau = 1$).
$\mathbf{p}$ is mapped to a one-hot vector using a function $\varphi(\cdot)$, such that $\varphi(\mathbf{p})_i = 1$ if $i = \argmax_j p_j$ and $0$ otherwise. 
As this function is non-differentiable, we must rely on the straight-through gradient estimator \parencite{bengio_estimating_2013} which assumes that the Jacobian ${\partial \, \varphi}/{\partial \, \mathbf{p}}$ equals the identity matrix. 
The one-hot vector can then be used for a codebook lookup to obtain the final quantized vector (e.g., $\mathbf{q}_t$ in \cref{eq_brief: w2v2 qtz}).

@@

\begin{sidewaystable*}
    \caption[Classification of selected latent variable models.]{ Selected latent variable models classified according the attributes defined throughout \cref{sec:plvms}. See \cref{tab:lvm-model-primitives} for the probability distributions that correspond to each of the attribute short-hands. \textbf{\textsc{hie}} indicates a hierarchical representation.}
    \label{tab:lvm-taxonomy}
    \begin{center}
        \renewcommand{\arraystretch}{1.1}
        \resizebox{0.99\textwidth}{!}{% <------ Don't forget this %
        \begin{tabular}{ l l | c c c | c c c c | c c c c c | c } 
            \toprule
            \multicolumn{2}{c}{} & 
            \multicolumn{3}{c}{\footnotesize\textsc{Observation}} & 
            \multicolumn{4}{c}{\footnotesize\textsc{Prior}} & 
            \multicolumn{5}{c}{\footnotesize\textsc{Inference}} \\
            \textbf{\textsc{model}} & 
            \textbf{\textsc{pub. date}} &
            \textbf{\textsc{arx}} & 
            \textbf{\textsc{loc}} & 
            \textbf{\textsc{glb}} &  
            \textbf{\textsc{arx}} & 
            \textbf{\textsc{arz}} & 
            \textbf{\textsc{ind}} & 
            \textbf{\textsc{glb}} &  
            \textbf{\textsc{arz}} & 
            \textbf{\textsc{flt}} & 
            \textbf{\textsc{lsm}} & 
            \textbf{\textsc{gsm}} & 
            \textbf{\textsc{glb}} &
            \textbf{\textsc{hie}} \\
            \midrule
            %                                                              OBSERVATION       |              PRIOR                |                  INFERENCE       
            %                                                         ARX      LOC      GLB      ARX      ARZ     IND      GLB      ARZ      FLT      LSM      GSM      GLB
            \textbf{VRNN} \footnotesize{\parencite{chung_recurrent_2015}}          & 2015 Jun. & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark \\
            \textbf{SRNN} \footnotesize{\parencite{fraccaro_sequential_2016}}      & 2016 May  & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark \\
            \textbf{HMM-VAE} \footnotesize{\parencite{ebbers_hidden_2017}}         & 2017 Mar. & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
            \textbf{ConvVAE} \footnotesize{\parencite{hsu_learning_2017}}          & 2017 Apr. & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \cmark & \xmark \\
            \textbf{FHVAE} \footnotesize{\parencite{hsu_unsupervised_2017}}        & 2017 Sep. & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \cmark & \cmark \\
            \textbf{VQ-VAE} \footnotesize{\parencite{oord_neural_2018}}            & 2017 Nov. & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\
            \textbf{BHMM-VAE} \footnotesize{\parencite{glarner_full_2018}}         & 2018 Mar. & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark \\
            \textbf{STCN} \footnotesize{\parencite{aksan_stcn_2019}}               & 2019 Feb. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
            \textbf{FDMM} \footnotesize{\parencite{khurana_factorial_2019}}        & 2019 Oct. & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark \\
            \textbf{ConvDMM} \footnotesize{\parencite{khurana_convolutional_2020}} & 2020 Jun. & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark \\
            \bottomrule
        \end{tabular}
    \end{center}
\end{sidewaystable*}

>>>>

\section{Introduction}
Representation learning has shaped modern computer vision \parencite{simonyan_very_2014} and natural language processing \parencite{devlin_bert_2018}, and more recently speech processing has been subject to the same development \parencite{baevski_wav2vec_2020}. Representation learning has been defined as \emph{``learning representations of the data that make it easier to extract useful information when building classifiers or other predictors"} \parencite{bengio_representation_2013}.  
Unsupervised representation learning is concerned with learning useful representations without the use of human annotations. Usually, a model is first pre"=trained on a task where plenty of data is available. The model is then fine"=tuned, or used to extract input representations for a smaller model, targeting a task with limited training data. In computer vision, both supervised \parencite{simonyan_very_2014, szegedy_going_2015, he_deep_2015} and unsupervised \parencite{pathak_context_2016, doersch_unsupervised_2015} representation learning have gained attention with supervised representation learning driven by the availability of large annotated datasets \parencite{deng_imagenet_2009}. For text and speech, pre"=training is usually unsupervised as labeled data is difficult to obtain. Although work on text has paved the way, and the two fields share many characteristics, learning representations from speech is a problem faced with a unique set of challenges.

@@

\begin{figure}
    \begin{flushleft}
        \textsc{\small \hspace{0.16\textwidth} self"=supervised \hspace{0.22\textwidth} latent variable}
    \end{flushleft}
    \centering
    \includegraphics[width=0.90\textwidth]{paper_brief/SSL_LVM.pdf}
    \caption[ Schematic overview of the self"=supervised and probabilistic latent variable models covered in this survey.]{A schematic overview of the two groups of models covered in this survey. \textit{Left}: A model trained with self"=supervised learning. We take these models to consist of two functions $f(\cdot)$ and $g(\cdot)$ (\cref{ssec:notation}). After pre"=training, $f(\cdot)$ is fine"=tuned or used for extracting features $\mathbf{c}$. $g(\cdot)$ is an auxiliary function used to accommodate the self"=supervised pre"=training task. \textit{Right}: A probabilistic latent variable model. In contrast to the self"=supervised model, the functions $f(\cdot)$ and $g(\cdot)$ learn the parameters of distributions $q$ and $p$. The latent variable $\mathbf{z}$ is commonly used for representation learning.}
    \label{fig:ssl_lvm}
\end{figure}

@@

When discussing self"=supervised models, we use $\mathbf{c}_{1:T}$ to denote a contextualized representation. For stochastic latent variable models, we use $\mathbf{z}_{1:T}$ as is customary to the field. 
While some models are frozen and produce representations used as input for downstream tasks (\textbf{\textsc{frz}}, \cref{tab:model-taxonomy}), others are designed to be fine"=tuned (\textbf{\textsc{ftn}}, \cref{tab:model-taxonomy}). In either case, we use $f(\cdot)$ to denote the model that is used for the downstream task. We use $g(\cdot)$ to denote any auxiliary model components (e.g., for a reconstruction task we might have $g: \mathbf{c}_t \mapsto \hat{\mathbf{x}}_t$). When a model can be naturally subdivided into multiple components, we simply use $f_*(\cdot)$ where $*$ may be any convenient descriptor. Finally, we often use a subscript when defining a loss, $\mathcal{L}_i$, to imply that the total loss is computed as a sum over $i$.

@@

Combining masking with a contrastive loss, wav2vec 2.0 was the first work to show that a competitive speech recognition model can be learned by fine"=tuning a pre"=trained model with as little as 10 minutes of labeled data. For this model
%
\begin{align}
    \mathbf{v}_{t} &= f_v(\mathbf{x}_{t-r:t+r}) \\
    \mathbf{c}_{1:T} &= f_c(\mathbf{v}_{1:T} \circ \mathbf{m}_{1:T}) \\
    \mathbf{q}_t &= g_q(\mathbf{v}_t) \enspace . \label{eq_brief: w2v2 qtz}
\end{align}

@@

\begin{table}
    \caption[Classification of selected self"=supervised and probabilistic latent variable models.]{ Selected models classified according to the binary attributes identified throughout the text. The models are sorted according to first publication date on arXiv which might differ from the citation year. \textbf{\textsc{msk}}: masking, \textbf{\textsc{prd}}: prediction, \textbf{\textsc{con}}: contrastive, \textbf{\textsc{rec}}: reconstruction, \textbf{\textsc{qtz}}: quantization, \textbf{\textsc{gen}}: generative, \textbf{\textsc{frz}}: frozen, \textbf{\textsc{ftn}}: fine"=tuned, \textbf{\textsc{loc}}: local, \textbf{\textsc{glo}}: global.}
    \label{tab:model-taxonomy}
    \begin{center}
        \setlength{\tabcolsep}{3pt}
        \renewcommand{\arraystretch}{1.1}
        \resizebox{1\textwidth}{!}{%
        \begin{tabular}{ l l | c c c c c c | c c c | c c } 
            \toprule

            \multicolumn{2}{c}{} & 
            \multicolumn{6}{c}{\textsc{Model and task design}} & 
            \multicolumn{3}{c}{\textsc{Resolution}} &
            \multicolumn{2}{c}{\makebox[0pt][c]{\textsc{Usage}}} \\
            \textbf{\textsc{model}} &
            \textbf{\textsc{pub. date}} &
            \textbf{\textsc{msk}} & 
            \textbf{\textsc{prd}} & 
            \textbf{\textsc{con}} &  
            \textbf{\textsc{rec}} &  
            \textbf{\textsc{qtz}} & 
            \textbf{\textsc{gen}} & 
            \textbf{\textsc{loc}} & 
            \textbf{\textsc{glb}} & 
            \textbf{\textsc{var}} & 
            \textbf{\textsc{frz}} & 
            \textbf{\textsc{ftn}} \\

@@

            \textbf{APC} \footnotesize{\parencite{chung_unsupervised_2019}}          & 2019 Oct. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 5/4
            \textbf{wav2vec} \footnotesize{\parencite{schneider_wav2vec_2019}}       & 2019 Apr. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/4
            \textbf{Mockingjay}  \footnotesize{\parencite{liu_mockingjay_2020}}      & 2019 Oct. & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark \\ % 25/10-19
            \textbf{wav2vec 2.0} \footnotesize{\parencite{baevski_wav2vec_2020}}     & 2020 Jun. & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 20/6
            \textbf{NPC} \footnotesize{\parencite{liu_nonautoregressive_2020}}       & 2020 Nov. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 1/11
            \textbf{DeCoAR 2.0} \footnotesize{\parencite{ling_decoar_2020}}          & 2020 Dec. & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\ % 11/12
            \textbf{SCPC} \footnotesize{\parencite{bhati_segmental_2021}}            & 2021 Jun. & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark \\ % 3/6
            \textbf{HuBERT} \footnotesize{\parencite{hsu_hubert_2021}}               & 2021 Jun. & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\ % 14/6
            \midrule
            \multicolumn{12}{c}{\textsc{Probabilistic latent variable models}} \\
            \midrule
            %                                                           MSK      PRD      CON      REC      QTZ      GEN      LOC      GLO     VAR      FRZ      FTN
            \textbf{VRNN} \footnotesize{\parencite{chung_recurrent_2015}}          & 2015 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
            \textbf{SRNN} \footnotesize{\parencite{fraccaro_sequential_2016}}      & 2016 May  & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
            \textbf{HMM-VAE} \footnotesize{\parencite{ebbers_hidden_2017}}         & 2017 Mar. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
            \textbf{ConvVAE} \footnotesize{\parencite{hsu_learning_2017}}          & 2017 Apr. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark & \cmark & \xmark \\
            \textbf{FHVAE} \footnotesize{\parencite{hsu_unsupervised_2017}}        & 2017 Sep. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \cmark & \xmark \\
            \textbf{VQ-VAE} \footnotesize{\parencite{oord_neural_2018}}            & 2017 Nov. & \xmark & \xmark & \xmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
            \textbf{BHMM-VAE} \footnotesize{\parencite{glarner_full_2018}}         & 2018 Mar. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
            \textbf{STCN} \footnotesize{\parencite{aksan_stcn_2019}}               & 2019 Feb. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
            \textbf{FDMM} \footnotesize{\parencite{khurana_factorial_2019}}        & 2019 Oct. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \cmark & \xmark \\
            \textbf{ConvDMM} \footnotesize{\parencite{khurana_convolutional_2020}} & 2020 Jun. & \xmark & \xmark & \xmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
            \bottomrule
        \end{tabular}
        }
    \end{center}
\end{table}

@@

\textit{Gumbel-softmax approach:} 
Say we want to quantize a vector $\mathbf{v}$ such that it takes one of $K$ possible values. We first map $\mathbf{v}$ to $\mathbf{l} \in \mathbb{R}^K$ and then map $\mathbf{l}$ to a probability vector $\mathbf{p} \in \mathbb{R}^K$ via the Gumbel softmax given by
\begin{align}
    p_i &= \frac{\exp(l_i + n_i) / \tau}{\sum_k^K \exp(l_k + n_k) / \tau} 
\end{align}
for $i=1,\dots,K$. Here $\tau$ is a temperature parameter and $\mathbf{n}\in\mathbb R^K$ is a random vector with $n_i = -\log(-\log(u_i))$ for $u_i \sim \mathcal{U}(0, 1)$.
As  $\tau \rightarrow 0$, $\mathbf{p}$ approaches a one-hot vector. The Gumbel noise $\mathbf{n}$ is a practical way to sample from the untempered categorical distribution (i.e., $\tau = 1$).
$\mathbf{p}$ is mapped to a one-hot vector using a function $\varphi(\cdot)$, such that $\varphi(\mathbf{p})_i = 1$ if $i = \argmax_j p_j$ and $0$ otherwise. 
As this function is non-differentiable, we must rely on the straight-through gradient estimator \parencite{bengio_estimating_2013} which assumes that the Jacobian ${\partial \varphi}/{\partial \mathbf{p}}$ equals the identity matrix. 
The one-hot vector can then be used for a codebook lookup to obtain the final quantized vector (e.g., $\mathbf{q}_t$ in \cref{eq_brief: w2v2 qtz}).

@@

\begin{table*}
    \caption[Classification of selected latent variable models.]{ Selected latent variable models classified according the attributes defined throughout \cref{sec:plvms}. See \cref{tab:lvm-model-primitives} for the probability distributions that correspond to each of the attribute short-hands. \textbf{\textsc{hie}} indicates a hierarchical representation.}
    \label{tab:lvm-taxonomy}
    \begin{center}
        \setlength{\tabcolsep}{3pt}
        \renewcommand{\arraystretch}{1.1}
        \resizebox{1\textwidth}{!}{%
        \begin{tabular}{ l l | c c c | c c c c | c c c c c | c } 
            \toprule
            \multicolumn{2}{c}{} & 
            \multicolumn{3}{c}{\textsc{Observation}} & 
            \multicolumn{4}{c}{\textsc{Prior}} & 
            \multicolumn{5}{c}{\textsc{Inference}} \\
            \textbf{\textsc{model}} & 
            \textbf{\textsc{pub. date}} &
            \textbf{\textsc{arx}} & 
            \textbf{\textsc{loc}} & 
            \textbf{\textsc{glb}} &  
            \textbf{\textsc{arx}} & 
            \textbf{\textsc{arz}} & 
            \textbf{\textsc{ind}} & 
            \textbf{\textsc{glb}} &  
            \textbf{\textsc{arz}} & 
            \textbf{\textsc{flt}} & 
            \textbf{\textsc{lsm}} & 
            \textbf{\textsc{gsm}} & 
            \textbf{\textsc{glb}} &
            \textbf{\textsc{hie}} \\
            \midrule
            %                                                              OBSERVATION       |              PRIOR                |                  INFERENCE       
            %                                                         ARX      LOC      GLB      ARX      ARZ     IND      GLB      ARZ      FLT      LSM      GSM      GLB
            \textbf{VRNN} \footnotesize{\parencite{chung_recurrent_2015}}          & 2015 Jun. & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark \\
            \textbf{SRNN} \footnotesize{\parencite{fraccaro_sequential_2016}}      & 2016 May  & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark \\
            \textbf{HMM-VAE} \footnotesize{\parencite{ebbers_hidden_2017}}         & 2017 Mar. & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
            \textbf{ConvVAE} \footnotesize{\parencite{hsu_learning_2017}}          & 2017 Apr. & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \cmark & \xmark \\
            \textbf{FHVAE} \footnotesize{\parencite{hsu_unsupervised_2017}}        & 2017 Sep. & \xmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \cmark & \cmark \\
            \textbf{VQ-VAE} \footnotesize{\parencite{oord_neural_2018}}            & 2017 Nov. & \cmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark \\
            \textbf{BHMM-VAE} \footnotesize{\parencite{glarner_full_2018}}         & 2018 Mar. & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark \\
            \textbf{STCN} \footnotesize{\parencite{aksan_stcn_2019}}               & 2019 Feb. & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark \\
            \textbf{FDMM} \footnotesize{\parencite{khurana_factorial_2019}}        & 2019 Oct. & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \cmark & \cmark & \cmark & \xmark & \xmark & \cmark & \cmark \\
            \textbf{ConvDMM} \footnotesize{\parencite{khurana_convolutional_2020}} & 2020 Jun. & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark \\
            \bottomrule
        \end{tabular}
        }
    \end{center}
\end{table*}
